Compositional Quantitative Resource Analysis of C Programs

### 2014-06-15 paper/main.tex
more work on section 2

\paragraph{Automatic Amortized Analysis}

By fixing the shape of a potential function $\Phi$, it becomes
possible to leverage existing tools to help finding a pair of
potential functions for an arbitrary program.  This is explained in
further details in the following sections.

We say that the analysis is amortized because the use of a potential
function $\Phi$ gives the flexibility to amortize some operations.
For example, if program states are simple mappings from variables to
natural numbers, we have the following judgement (provided that
initially $\state(\code x) \ge \state(\code y) \ge 0$).
$$
\htriple
{\lambda \state . \, 2 {\cdot} \state (\code x) + 10}
{\code{ x = x -  y}}
{\lambda \state . \, 2 {\cdot} \state(\code x) + 2 {\cdot} \state(\code y) + 10}
$$
This must be understood as a motion of potential from the
variable~$\code x$ to the variable~$\code y$. If some resource is
consumed $\code y$ times afterwards, we can then use $\code y$'s
potential to pay for it. However, the initial potential function
$\Phi$ only refers to the variable~$\code x$: the potential function
allows us to pass on the cost of the subsequent resource consumption
to~$\code x$.  This flexibility combined with intelligent decisions on
potential motions allow our tool to discover linear invariants for
surprisingly convoluted programs.

>>>>

We have implemented and proved sound such a quantitative Hoare for
CompCert Clight in the Coq Proof Assistant.  Because we use a shallow
embedding in Coq, every function that can be expressed in Coq can in
principle be used as a resource bound.  We can use the logic to prove
a concrete bound for a given resource metric, or a bound that is
parametric in set of resource metrics.  It is also possible to
incorporate boolean preconditions into the bounds to express that the
bound is only valid for a certain class of inputs.  To this end, we
allow the potential function $\Phi$ in the pre and post conditions to
take non-negative number or $\infty$ (infinity) as values.  Infinity,
plays the same role as $\bot$ in Hoare logic.  Boolean assertions can
be embedded into potential functions by mapping $\bot$ to $\infty$ and
$\top$ to $0$.

It is for instance possible to derive the following quantitative
triple that holds for all stack metrics $M$.  A stack metric assigns
cost $0$ to all evaluation steps except function calls.  Before a call
$f(x)$, we consume $M(f) > 0$ resources and after the call we return
$-M(f)$ resources.  So $M(f)$ corresponds to the stack-frame size of
the function $f$.
$$
XXX
$$
XXX explanation of triple XXX

We say that the analysis is amortized because the use of a potential
function $\Phi$ gives the flexibility to amortize some operations.
For example, if program states are simple mappings from variables to
natural numbers, we have the following judgement (provided that
initially $\state(\code x) \ge \state(\code y) \ge 0$).
$$
\htriple
{\lambda \state . \, 2 {\cdot} \state (\code x) + 10}
{\code{ x = x -  y}}
{\lambda \state . \, 2 {\cdot} \state(\code x) + 2 {\cdot} \state(\code y) + 10}
$$
This must be understood as a motion of potential from the
variable~$\code x$ to the variable~$\code y$. If some resource is
consumed $\code y$ times afterwards, we can then use $\code y$'s
potential to pay for it. However, the initial potential function
$\Phi$ only refers to the variable~$\code x$: the potential function
allows us to pass on the cost of the subsequent resource consumption
to~$\code x$.  This flexibility combined with intelligent decisions on
potential motions allows us to design an automatic analysis to
discover linear invariants for surprisingly convoluted programs.

\subsection{Automatic Amortized Analysis}

A program logic provides a principled foundation for statically
analyzing programs.  However, program logics need to be supported by
automatic methods to be useful practice.  That is why we developed a
new automatic amortized analysis for deriving quantitative Hoare
triples for common C programs.

To this end, we fix the shape of the potential functions $\Phi$ so
that it becomes possible to use \emph{linear programming} to compute a
derivation in the logic.  If we assume that a program state $\state$
simply maps variable to integers then we require
$$
\Phi(\state) = \sum_{x,y \in \dom{\state}} q_{(x,y)} {\cdot} |[\state(x),\state(y)]|
$$
where $q_{(x,y)} \in \Qplus$ and $|[a,b]| = \max(b-a,0)$.
